---
title: "Preparing data and integration points"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Preparing data and integration points}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

Every mlumr analysis follows the same four-step pipeline. This vignette covers
the **family-agnostic** machinery, combining IPD and AgD and, above all, the
numerical integration that powers population adjustment. The exact `set_ipd()` /
`set_agd()` arguments for each outcome family are shown, with complete runnable
examples, in the per-outcome vignettes (see the reference table below).

1. `set_ipd()`, individual patient data (the index treatment)
2. `set_agd()` (or `set_agd_surv()` for survival), aggregate data (comparator)
3. `combine_data()`, join them into one `mlumr_data` object
4. `add_integration()`, quasi-Monte Carlo integration points (ML-UMR only)

## The pipeline (illustrated on a binary endpoint)

The four steps below build a complete `mlumr_data` object from a binary IPD trial
and a one-row aggregate comparator; the sections that follow explain each piece
(covariate naming, integration, correlation) in detail:

```{r, eval = TRUE}
library(mlumr)
set.seed(2026)

# IPD (index): one row per patient
trial_a <- data.frame(
  treatment = "Drug_A",
  response = rbinom(500, 1, 0.6),
  age_group = rbinom(500, 1, 0.4),
  sex = rbinom(500, 1, 0.55)
)
ipd <- set_ipd(trial_a, treatment = "treatment", outcome = "response",
               covariates = c("age_group", "sex"))

# AgD (comparator): one row per study, with covariate summaries
trial_b <- data.frame(
  treatment = "Drug_B", n_total = 400, n_events = 160,
  age_group_mean = 0.35, sex_prop = 0.50
)
agd <- set_agd(trial_b, treatment = "treatment",
               outcome_n = "n_total", outcome_r = "n_events",
               cov_means = c("age_group_mean", "sex_prop"),
               cov_types = c("binary", "binary"))

dat <- combine_data(ipd, agd)
dat
```

## Outcome-family input reference

The outcome arguments differ by family; everything else (covariates, combining,
integration) is identical. Each family has a complete worked example.

| Family | `set_ipd()` outcome args | `set_agd()` outcome args | Vignette |
|--------|--------------------------|--------------------------|----------|
| binomial | `outcome` (0/1) | `outcome_n`, `outcome_r` | `vignette("binary-outcomes")` |
| normal | `outcome`, `family = "normal"` | `outcome_mean`, `outcome_se` (+ `outcome_n`) | `vignette("continuous-outcomes")` |
| poisson | `outcome`, `exposure`, `family = "poisson"` | `outcome_r`, `outcome_E` | `vignette("count-outcomes")` |
| survival | `Surv`/`time`/`status`, `family = "survival"` | `set_agd_surv()`: pseudo-IPD `time`/`status` | `vignette("survival-outcomes")` |

`set_ipd()` validates that the outcome matches the family (binary 0/1; numeric
for normal; non-negative integer counts for Poisson; positive times for
survival) and drops rows with missing values (with a warning).

### Covariate naming and types

`set_agd()` strips `_mean` and `_prop` suffixes so AgD covariate names match the
IPD (`age_group_mean` → `age_group`, `sex_prop` → `sex`). Declare each covariate
as `"binary"` or `"continuous"` via `cov_types`; continuous covariates also need
a standard-deviation column in `cov_sds`.

### Scale conventions for aggregate summaries

For the **normal** family, `outcome_mean` and `outcome_se` must be on the
**original (arithmetic) scale**, the same scale as the IPD outcome, even when
fitting with `link = "log"`. Back-transform any geometric mean or log-scale
summary (propagating the SE by the delta method) before calling `set_agd()`.
See `vignette("continuous-outcomes")`.

## Integration points

ML-UMR adjusts for covariate imbalance by integrating the IPD-derived outcome
model over the comparator population's covariate distribution. mlumr represents
that distribution by **Sobol quasi-Monte Carlo (QMC)** integration points,
correlated with a **Gaussian copula** so the joint covariate structure is
respected.

```{r, eval = TRUE}
dat <- add_integration(
  dat,
  n_int = 64,
  age_group = distr(qbern, prob = age_group_mean),
  sex = distr(qbern, prob = sex_mean)
)
```

- **`n_int`**, number of points (powers of 2: 32, 64, 128, …). More points give
  better accuracy at the cost of slower fitting; check adequacy with
  `check_integration()` (below).
- **`distr()`**, wraps a quantile function whose parameters reference AgD
  summary columns. Use `qbern` (Bernoulli, supplied by mlumr) for binary
  covariates, `qnorm` for normal, `qgamma` for skewed positive covariates, and
  `qlogitnorm` for a covariate reported as a proportion on `(0, 1)`.

mlumr's `qgamma()` and `qlogitnorm()` accept a `mean` and `sd` on the natural
scale in place of their native parameters, so the moments can be taken from a
published baseline table without conversion. Both must be supplied together.
A logit-normal additionally requires `sd^2 < mean * (1 - mean)`, the bound any
variable on `(0, 1)` satisfies; a percentage must be rescaled to a proportion
first.

```{r}
add_integration(
  dat, n_int = 64,
  age = distr(qbern, prob = age_mean),                  # binary
  bmi = distr(qnorm, mean = bmi_mean, sd = bmi_sd),     # continuous (normal)
  biomarker = distr(qgamma, mean = bio_mean, sd = bio_sd),  # right-skewed
  bsa = distr(qlogitnorm, mean = bsa_mean, sd = bsa_sd)     # proportion on (0, 1)
)
```

### Correlation handling

By default the covariate correlation is estimated from the IPD (Spearman) and
adjusted for the Gaussian copula. For binary and mixed discrete/continuous
pairs this copula adjustment is a **pragmatic approximation**: the exact latent
Gaussian correlation depends on the marginal prevalences, which the current
transform does not use. It is adequate for most datasets, but if you have strong
prior information, or covariates with prevalences near 0/1, strong
correlations, or many covariates, supply your own matrix via `cor` and confirm
the result is stable with `check_integration()`.

```{r}
my_cor <- matrix(c(1, 0.3, 0.3, 1), 2, 2)
add_integration(dat, n_int = 64, cor = my_cor,
                age_group = distr(qbern, prob = age_group_mean),
                sex = distr(qbern, prob = sex_mean))

add_integration(dat, n_int = 64, cor_adjust = "none",  # no adjustment
                age_group = distr(qbern, prob = age_group_mean),
                sex = distr(qbern, prob = sex_mean))
```

### Inspecting the integration points

`unnest_integration()` expands the stored points to long format, one row per
integration point per covariate, so you can inspect or plot the realized
comparator distribution directly:

```{r, eval = TRUE}
head(unnest_integration(dat))     # long format
```

```{r, eval = TRUE, purl = FALSE}
check_integration(
  dat,
  age_group = distr(qbern, prob = age_group_mean),
  sex = distr(qbern, prob = sex_mean)
)
```

`check_integration()` compares the realized integration points against the
requested marginals and correlations and flags when `n_int` is too small, run
it whenever there are several or strongly correlated covariates.

### Visualizing the integration points

For a continuous covariate the quasi-Monte Carlo points should reproduce the
requested marginal. Here we build a small example with a continuous `age` and
overlay the integration points on the requested Normal density:

```{r viz-int, eval = TRUE, fig.width = 7, fig.height = 4, fig.align = "center"}
set.seed(2026)
ipd_c <- data.frame(treatment = "Drug_A", response = rbinom(400, 1, 0.5),
                    age = rnorm(400, 55, 10))
agd_c <- data.frame(treatment = "Drug_B", n_total = 300, n_events = 120,
                    age_mean = 60, age_sd = 9)
dat_c <- combine_data(
  set_ipd(ipd_c, treatment = "treatment", outcome = "response", covariates = "age"),
  set_agd(agd_c, treatment = "treatment", outcome_n = "n_total",
          outcome_r = "n_events", cov_means = "age_mean", cov_sds = "age_sd",
          cov_types = "continuous"))
dat_c <- add_integration(dat_c, n_int = 128,
                         age = distr(qnorm, mean = age_mean, sd = age_sd))
pts <- unnest_integration(dat_c)

if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  ggplot(pts, aes(age)) +
    geom_histogram(aes(y = after_stat(density)), bins = 25,
                   fill = "#3B6B9A", alpha = 0.5) +
    stat_function(fun = dnorm, args = list(mean = 60, sd = 9),
                  color = "darkred", linewidth = 0.8) +
    labs(x = "Age (comparator integration points)", y = "Density",
         title = "QMC integration points reproduce the requested Normal(60, 9)") +
    theme_minimal()
}
```

## Integration requirements for STC and naive

`naive()` can be called immediately after `combine_data()` because it uses only
the observed crude outcomes. `stc()` requires integration points for nonlinear
links, including binomial and Poisson outcomes and normal-log models. Replacing
a covariate distribution by its mean before applying a nonlinear inverse link
does not recover the marginal outcome. Only the normal identity-link STC can use
the aggregate means directly; see `vignette("choosing-a-method")`.

```{r, eval = TRUE}
dat_no_int <- combine_data(ipd, agd)
naive(dat_no_int)
stc(dat)
```
